---
title: "Homework 8 - AGST 5014"
author: "Igor Kuivjogi Fernandes and Ashmita Upadhyay"
date: "`r Sys.Date()`"
output: pdf_document
---

__1. The following experiment comes from a central composite design with 4 factors in 2 blocks. Conduct the proper analysis (including graphs, interpretation, etc). __ 

```{r}
q1 <- read.csv("HW8_Q1.csv")
q1 <- transform(q1, block = factor(block), logSD = NULL)
str(q1)
```

```{r}
table(q1$block)
```

We have 18 experimental units for the factorial points, and 12 for the axial points.   

Let's fit some Response Surface Methodology models.

First, we can start with a simple first order model:
```{r}
library(rsm)

mod1 <- rsm(ave ~ block + FO(x1, x2, x3, x4), data = q1)
summary(mod1)
```

The lack of fit is significant (p-value $< 0.05$), so we should include more complex terms.  

Now a second-order model:
```{r}
mod2 <- rsm(ave ~ block + SO(x1, x2, x3, x4), data = q1)
summary(mod2)
```

There are positive and negatives eigenvalues, which means we have a saddle point.   
The $\text{adj.} R^2 = 0.9078$, and the lack of fit is not significant, so we can stick with this model. 

The optimal experimental points are:
```{r}
summary(mod2)$canonical$xs
```


```{r, fig.asp = 1.2}
par(mfrow = c(3, 2))
contour(
  mod2,
  ~ x1 + x2 + x3 + x4,
  image = TRUE
)
```

We have 6 plots because there are 4 two-way interactions (`x1:x2`, `x1:x3`, `x1:x4`, `x2:x3`, `x2:x4`, and `x3:x4`).   
From the plots we can see that the maximum point reached lies roughly in the interval 380-390.   
We can check the distribution of the fitted values:   
```{r}
summary(mod2$fitted.values)
```

The lowest point was 346.8, whereas the maximum point was 397.1.   

Now we can run a steepest-ascent algorithm to search for a better solution:
```{r}
steep <- steepest(mod2)
steep
```

The optimal solution points now are:
```{r}
opt_points <- steep[which.max(steep$yhat), ]
opt_points
```

These are the new points we would use in the process to get the higher response values.      
If we do predictions using these new coefficients, we get the maximum predicted response found by the steepest-ascent algorithm above:
```{r}
grid <- expand.grid(
  block = unique(q1$block), 
  x1 = opt_points$x1,
  x2 = opt_points$x2,
  x3 = opt_points$x3,
  x4 = opt_points$x4
)
predict(mod2, grid)
```

As we have 2 blocks, the model did two predictions. The first matches with the steepest-ascent algorithm.   

__2. The design below presents the yield of different common bean cultivars. There was a variable stand count in each plot. Conduct the proper analysis.__ 

```{r}
q2 <- read.csv("HW8_Q2.csv")
q2 <- transform(q2, block = factor(block), cv = factor(cv))
str(q2)
```

```{r}
table(q2$block)
```

The blocks are equally frequent.   

The `stand` variable is a covariable, hence we can use ANCOVA to analyse the yield of different cultivars.    

First step is to check whether `stand` is independent of the treatment `cv`.   
```{r}
check <- aov(stand ~ block + cv, data = q2)
summary(check)
```

The `cv` is not significant, so we expect `stand` and `cv` to not be related.   

Next, we run ANCOVA with interaction. For ANCOVA, we should use Type III sum of squares.   
```{r}
check_inter <- lm(
  yield ~ block + cv * stand, 
  contrasts = list(cv = contr.sum), 
  data = q2
)
car::Anova(check_inter, type = 'III')
```

The interaction `cv:stand` is not significant, so we can go further and fit ANCOVA without the interaction: 
```{r}
ancova <- lm(
  yield ~ block + cv + stand, 
  contrasts = list(cv = contr.sum),
  data = q2
)
car::Anova(ancova, type = 'III')
```

The `cv` is indeed significant, using a significance level of $\alpha = 0.05$.   

Now let's check the usual ANOVA assumptions:
```{r, fig.asp = 1.2}
par(mfrow= c(2, 1))
plot(ancova, which = 1)
plot(ancova, which = 2)
```

We have homogeneous variance across the fitted values and the residuals seems to be normally distributed.

Which cultivar was the best?
```{r}
plot(
  emmeans::emmeans(ancova, pairwise ~ cv, adjust = 'tukey'), 
  interval = F, comparisons = T
)
```

We can see that cultivar `GBrilhante` had better performance than `CNFP8022`, but `GBrilhante` is not different from others.         

__3. Design a proper experiment to identify the best dose of Nitrogen and amount of water to maximize yield (choose what values you would use). __

#### Nitrogen:   
- Levels: $100$ and $200$    
- Center: $\frac{100 + 200}{2} = 150$    
- Range = $200 - 100 = 100$    

#### Water:   
- Levels: $10$ and $30$
- Center: $\frac{10 + 30}{2} = 20$
- Range: $30 - 10 = 20$

```{r}
nitro_levels <- c(100, 200)
nitro_center <- mean(nitro_levels)
nitro_range <- max(nitro_levels) - min(nitro_levels)
nitro_center; nitro_range
```


```{r}
water_levels <- c(10, 30)
water_center <- mean(water_levels)
water_range <- max(water_levels) - min(water_levels)
water_center; water_range
```


```{r}
set.seed(2023)
q3 <- ccd(
    basis = 2,  # 2 factors (nitrogen and water)
    n0 = c(2, 2),  # 2 central point reps for each block 
    alpha = 'rotatable',
    coding = list(
      x1 ~ (nitrogen - nitro_center) / (0.5 * nitro_range),
      x2 ~ (water - water_center) / (0.5 * water_range)
    )
)
q3
```

Convert to data.frame and generate yield:
```{r}
set.seed(2023)
q3y <- as.data.frame(q3)
q3y$yield <- rnorm(nrow(q3y), 10, 1) 
q3y[(q3y$x1 == 0) & (q3y$x2 <= 1.3), 'yield'] <- (
  q3y[(q3y$x1 == 0) & (q3y$x2 <= 1.3), 'yield'] + rnorm(5, 4, 1)
)
q3y
```

Now we can fit Response Surface Methodology models:

First, a first-order model:   
```{r}
mod1 <- rsm(yield ~ FO(x1, x2), data = q3y)
summary(mod1)
```

Despite we don't have a significant lack of fit (at significance level of $\alpha = 0.05$), we have a bad adj. $R^2$.   
We can try a more complex model such as second-order model:
```{r}
mod2 <- rsm(yield ~ SO(x1, x2), data = q3y)
summary(mod2)
```
This model is better, because the adj. $R^2 \approx 51$ and the lack of fit is not significant.   
The negatives eigenvalues indicate a maximum point.  

The optimal experimental points are:
```{r}
opt_exp_points <- summary(mod2)$canonical$xs
opt_exp_points
```

In the original scale we have to back transform the coded variables:
```{r}
opt_orig_points <- c()
opt_orig_points[1] <- ((0.5 * nitro_range) * opt_exp_points[1]) + nitro_center
opt_orig_points[2] <- ((0.5 * water_range) * opt_exp_points[2]) + water_center
names(opt_orig_points) <- c('nitrogen', 'water')
opt_orig_points
```


```{r}
contour(
  mod2,
  ~ x1 + x2,
  image = TRUE
)
```

The maximum seems to be close to the range $x1 = 0$ and $x2 < 0$.     
We can check the distribution of the fitted values:   
```{r}
summary(mod2$fitted.values)
```

The maximum yield found was 13.978.   

Maybe we can find a highest point using steepest-ascent method:
```{r}
steep <- steepest(mod2)
steep
```

Maximum point:
```{r}
df_opt_points <- steep[which.max(steep$yhat), ]
df_opt_points
```

The maximum yield was found using $x = 0.051$ and $x2 = -0.497$.    
If we do a predict using these points we can reach to the same maximum yield:   
```{r}
predict(mod2, df_opt_points[, c('x1', 'x2')])
```

We can back transform and check what are the best treatment levels to reach the optimum yield:
```{r}
opt_points <- c()
opt_points[1] <- ((0.5 * nitro_range) * df_opt_points$x1) + nitro_center
opt_points[2] <- ((0.5 * water_range) * df_opt_points$x2) + water_center
names(opt_points) <- c('nitrogen', 'water')
opt_points
```


<!-- Let's suppose there are 3 different doses of nitrogen and 2 different levels of water.    -->
<!-- We can design a CRD with 2 replications:    -->
<!-- ```{r} -->
<!-- set.seed(2023) -->

<!-- # create design -->
<!-- q3 <- expand.grid( -->
<!--   nitrogen = c(10, 50, 150),  -->
<!--   water = c(0, 20) -->
<!-- ) -->
<!-- q3 <- rbind(q3, q3)  # 2 reps -->
<!-- q3 <- q3[sample(1:nrow(q3)), ] -->
<!-- rownames(q3) <- 1:nrow(q3) -->
<!-- q3 <- transform(q3, nitrogen = factor(nitrogen), water = factor(water)) -->

<!-- # simulate yield -->
<!-- q3$yield <- rnorm(nrow(q3), 6, 1) -->

<!-- q3[(q3$nitrogen == 10) & (q3$water == 0), 'yield'] <- q3[(q3$nitrogen == 10) & (q3$water == 0), 'yield'] - rnorm(2, 5, 1) -->
<!-- q3[(q3$nitrogen == 10) & (q3$water == 20), 'yield'] <- q3[(q3$nitrogen == 10) & (q3$water == 20), 'yield'] + rnorm(2, 2, 1) -->

<!-- q3[(q3$nitrogen == 50) & (q3$water == 0), 'yield'] <- q3[(q3$nitrogen == 50) & (q3$water == 0), 'yield'] + rnorm(2, 2, 1) -->
<!-- q3[(q3$nitrogen == 50) & (q3$water == 20), 'yield'] <- q3[(q3$nitrogen == 50) & (q3$water == 20), 'yield'] - rnorm(2, 3, 1) -->

<!-- q3[(q3$nitrogen == 150) & (q3$water == 0), 'yield'] <- q3[(q3$nitrogen == 150) & (q3$water == 0), 'yield'] + rnorm(2, 1, 1) -->
<!-- q3[(q3$nitrogen == 150) & (q3$water == 20), 'yield'] <- q3[(q3$nitrogen == 150) & (q3$water == 20), 'yield'] + rnorm(2, 4, 1) -->

<!-- ``` -->

<!-- Let's check the yield distribution: -->
<!-- ```{r} -->
<!-- plot(yield ~ nitrogen, data = q3) -->
<!-- ``` -->

<!-- Seems larger doses of nitrogen increases yield. -->

<!-- ```{r} -->
<!-- plot(yield ~ water, data = q3) -->
<!-- ``` -->

<!-- Maybe larger amount of water also increases yield.    -->

<!-- ```{r} -->
<!-- with(q3, interaction.plot(nitrogen, water, yield)) -->
<!-- ``` -->

<!-- Seems there is an interaction.    -->

<!-- Let's run ANOVA: -->
<!-- ```{r} -->
<!-- mod <- aov(yield ~ nitrogen * water, data = q3) -->
<!-- summary(mod) -->
<!-- ``` -->

<!-- As we expect, all the terms are significant.    -->

<!-- ```{r, fig.asp = 1} -->
<!-- par(mfrow = c(2, 2)) -->
<!-- plot(mod, which = 5) -->
<!-- plot(mod, which = 1) -->
<!-- plot(mod, which = 2) -->
<!-- plot(mod$residuals ~ rownames(q3),  -->
<!--      main = 'Residuals vs Exp. Units', font.main = 1, -->
<!--      xlab= 'Experimental Unit', ylab = 'Residuals') -->
<!-- abline(h = 0, lty = 2, col = 'grey') -->
<!-- ``` -->

<!-- Seems all the residuals plots are good, therefore our model is adequate.    -->

<!-- Now that we have the significant terms, we can try to search for the best dose of nitrogen and amount of water that maximizes yield.     -->





