---
title: "Homework 7 - AGST 5014"
author: "Igor Kuivjogi Fernandes and Ashmita Upadhyay"
date: "`r Sys.Date()`"
output: pdf_document
---

__1. A soil scientist conducted an experiment to evaluate the effects of soil 
compaction and soil moisture on the activity of soil microbes. Reduced levels of
microbe activity will occur in poorly aerated soils. The aeration levels can be 
restricted in highly saturated or compacted soils. Treated soil samples were placed
in airtight containers and incubated under conditions conducive to microbial activity.
The microbe activity in each soil sample was measured as the percent increase in
CO2 produced above atmospheric levels. The treatment design was a 3x3 factorial
with three levels of soil compaction (bulk density = mg soil/m$^3$) and three levels
of soil moisture (kg water/kg soil). There were two replicate soil container units
prepared for each treatment. The CO2 evolution/kg soil/day was recorded on three
successive days. The data for each soil container unit are shown below:__    

```{r}
q1 <- read.csv('HW7_Q1.csv')
q1 <- transform(
  q1, Density = factor(Density), Moisture = factor(Moisture), Unit = factor(Unit)
)
str(q1)
```

a) Describe this experiment (treatment and experimental design) and write the 
statistical model.     

This is a Repeated Measures design, because the data for each soil container was collected on three successive days.   
We have two treatments:   
- soil compaction (density) with three levels (1.1, 1.4, 1.6)    
- soil moisture with three levels (0.1, 0.2, 0.24)   

```{r, echo = F, eval = F}
# from https://online.stat.psu.edu/stat510/lesson/10/10.1
# animal is random
# aov(Y ~ Treatment * Time + Error(Animal), phlebitisdata)
```

Statistical model: 
$$
y = \mu + \alpha + \beta + \tau + \alpha \tau + \beta \tau + \alpha \beta + \alpha \beta \tau + \varepsilon,
$$

where $y$ is the response, $\alpha$ is the density treatment, $\beta$ is the moisture treatment, $\tau$ is the day (time) of measurement, $\alpha \tau$ is the density interacting with time, $\beta \tau$ is the moisture interacting with time, $\alpha \beta$ in the interaction between density and moisture, $\alpha \beta \tau$ is the interaction between density, moisture and time, and $\varepsilon$ is the error term.    


b) Conduct the proper statistical analysis. Are the assumptions for this type of analysis met?   

Firstly, we can try a multivariate model and check the assumption of sphericity.      
```{r}
mod_multi <- lm(cbind(day1, day2, day3) ~ Density * Moisture, data = q1)
mod <- car::Anova(mod_multi, idata = data.frame(day = factor(1:3)), idesign = ~day)
summary(mod, multivariate = F)$sphericity.tests
```

For Mauchly Tests for Sphericity, we reject the null hypothesis that the covariance matrix of the repeated measures obeys the Huynh-Feldt condition. In this case, we cannot use the split-plot framework and it's better to stick with a model that is able to account for variance-covariance structures.      

Let's do some plots for checking the trend of response.
```{r, fig.asp = 1.3}
# wide to long format
q1_long <- q1 |>
  tidyr::pivot_longer(-c(Density, Moisture, Unit), names_to = 'Day', values_to = 'Y') |>
  dplyr::mutate(Day = factor(stringr::str_sub(Day, 4)))

par(mfrow = c(2, 1))
with(q1_long, interaction.plot(Day, Density, Y))
with(q1_long, interaction.plot(Day, Moisture, Y))
```

Seems the trend in response for density is quite the same for every density level, whereas for moisture the trends are different.    

Let's fit linear models with different variance-covariance structures.   
```{r, message = F}
library(nlme)

# identity
mod_id <- gls(
  Y ~ Density * Moisture * Day, 
  data = q1_long, 
)

# compound symmetry
mod_cs <- gls(
  Y ~ Density * Moisture * Day, 
  corr = corCompSymm(form = ~ 1 | Unit),  
  data = q1_long, 
)

# diagonal
mod_diag <- gls(
  Y ~ Density * Moisture * Day, 
  weights = varIdent(form = ~ 1 | Day),
  data = q1_long, 
)

# 1st order auto regressive
mod_ar1 <- gls(
  Y ~ Density * Moisture * Day, 
  corr = corAR1(form = ~ 1 | Unit),  
  data = q1_long, 
)

# unstructured
mod_us <- gls(
  Y ~ Density * Moisture * Day, 
  corr = corSymm(form = ~ 1 | Unit),  
  weights = varIdent(form = ~ 1 | Day),
  data = q1_long, 
)

# comparing models
anova(mod_id, mod_cs, mod_diag, mod_ar1, mod_us)
```

The AIC and BIC tell us that the model 4 (1st order auto regressive) is the best model. The LR tests also shows us that the model 4 is better than
model 3 (test 3 vs 4, p-value = $0.0001$) and model 5 (4 vs 5, p-value = $0.2054$). Hence, we chose model 4.
Let's check some assumptions:   
```{r}
plot(mod_ar1$fitted, mod_ar1$residuals)
```
Seems we have homogeneous variance while increasing the fitted values, which is good.   

```{r}
shapiro.test(mod_ar1$residuals)
```

We also don't reject the hypothesis that the residuals are normally distributed, using a significance level of $\alpha = 0.05$.   

```{r}
anova(mod_ar1)
```

Both `Density` and `Moisture` treatments are significant, as well the interaction `Moisture:Day`. The interactions `Density:Moisture`, `Density:Day` and `Density:Moisture:Day` are not significant, all using a significance level of $\alpha = 0.05$.     

__2. An agronomist conducted a yield trial with five alfalfa cultivars in a randomized
complete block design with three replications. Each plot was harvested four times
in each of two years. The plot yield (lb/plot) from two harvests from each plot in each of two years are shown in the table below. Compare the yield of the cultivars (notice that you need adjusted means for that).__    

```{r}
q2 <- read.csv('HW7_Q2.csv')
q2 <- transform(q2, Cultivar = factor(Cultivar), Block = factor(Block))
str(q2)
```


```{r}
# pivot to long with 4 measurements
# q2_long <- q2 |>
#   tidyr::pivot_longer(-c(Cultivar, Block), names_to = 'Date', values_to = 'Y') |>
#   dplyr::mutate(Date = factor(Date))
  
# pivot to long with 2 measurements
q2_long <- q2 |>
  tidyr::pivot_longer(-c(Cultivar, Block), names_to = 'Date', values_to = 'Y') |>
  dplyr::mutate(Year = factor(ifelse(Date %in% c('Apr.86', 'May.86'), 86, 87))) |>
  dplyr::mutate(Month = stringr::str_sub(Date, end = 3)) |>
  dplyr::select(-Date) |>
  tidyr::pivot_wider(id_cols = c(Cultivar, Block, Year), names_from = 'Month', values_from = 'Y') |>
  dplyr::arrange(Year, Cultivar, Block) |>
  tidyr::pivot_longer(c(Apr, May), names_to = 'Month', values_to = 'Y') |>
  dplyr::mutate(number = 1) |>
  dplyr::group_by(Cultivar, Block, Year) |>
  dplyr::mutate(Unit = cumsum(number))

q2_long
```

We could try include the year as a factor and assume there are only 2 measurements in time, but we can also analyse it with 4 measurements, although they are not evenly spaced. We stick with the latter.      

<!-- ```{r, message = F} -->
<!-- # identity -->
<!-- mod_id <- gls( -->
<!--   Y ~ Cultivar * Date,  -->
<!--   data = q2_long,  -->
<!-- ) -->

<!-- # compound symmetry -->
<!-- mod_cs <- update(mod_id, corr = corCompSymm(form = ~ 1 | Block)) -->

<!-- # diagonal -->
<!-- mod_diag <- update(mod_id, weights = varIdent(form = ~ 1 | Date)) -->

<!-- # 1st order auto regressive -->
<!-- mod_ar1 <- update(mod_id, corr = corAR1(form = ~ 1 | Block)) -->

<!-- # unstructured -->
<!-- # not converging -->
<!-- # mod_us <- update(mod_id, corr = corSymm(form = ~ 1 | Block), weights = varIdent(form = ~ 1 | Date)) -->

<!-- # comparing models -->
<!-- anova(mod_id, mod_cs, mod_diag, mod_ar1) -->
<!-- ``` -->

<!-- It's clear that identity model was the best (lower AIC, lower BIC, and LR test was not significant, i.e. model 1 is better than model 2).    -->

We can try to check sphericity and check whether a split-plot analysis would be adequate here.   
```{r}
mod_multi <- lm(cbind(Apr.86, May.86, Apr.87, May.87) ~ Cultivar, data = q2)
mod <- car::Anova(mod_multi, idata = data.frame(Date = factor(1:4)), idesign = ~Date)
summary(mod, multivariate = F)$sphericity.tests
```

We don't reject the null hypothesis that there is a departure from the Huynh-Feldt, using a significance level of $\alpha = 0.05$, which means we can use the split-plot analysis. In this case, the `Cultivar` is the whole-plot, and the `Date` is the sub-plot. We treat block as fixed.      
```{r}
mod_sp <- aov(Y ~ Cultivar * Date + Error(Block/Cultivar), data = q2_long)
summary(mod_sp)
# mm_sp <- lme(
#   fixed = Y ~ Cultivar * Date,
#   random = ~ 1 | Block/Cultivar,  # or Block + Block:Cultivar
#   method = 'REML',
#   data = q2_long
# )
# anova(mm_sp)
```

The error term `Block:Cultivar` error is being used to calculate the F-value of `Cultivar` (the whole-plot), which effect is not significant.      
The usual residual term is being used to calculate the F-value of `Date`, which is significant, i.e. the yield changes over time.     

Although the `Cultivar` is not significant, the change over time is significant, i.e. `Cultivar:Date`, who seems to make sense to compare this term to found which cultivar was better across the time.   
```{r}
emmeans::emmeans(mod_sp, pairwise ~ Cultivar, by = 'Date')$contrasts
```



The `Cultivar` is not significant, using a significance level of $\alpha = 0.05$. Hence, all the cultivars performed the same. 

```{r}
emmeans::
```


__3. One experiment was set up to assess possible phytotoxicity effects relating
to an excessive persistence of herbicide residues in soil. The three crops
were sown 40 days after a herbicide treatment (a check was included as the second
herbicide treatment).__    
```{r}
q3 <- read.csv("HW7_Q3.csv")
```

a) Describe this experiment and write the statistical model.



b) Run the proper analyze on experiment and proceed with 
interpretations. 



