---
title: "Homework 5 - AGST 5014"
author: "Igor Kuivjogi Fernandes and Ashmita Uphadhay"
date: "`r Sys.Date()`"
output: pdf_document
---


__1.	The following data set has 2 mistakes. Find the issue and analyze the experiment.__  
__Include all the necessary analysis and interpretation__
```{r}
q1 <- read.csv("HW5_Q1.csv")
q1$row <- as.factor(q1$row)
q1$col <- as.factor(q1$col)
q1$rep <- as.factor(q1$rep)
q1$cultivar <- as.factor(q1$cultivar)
q1$fertilizer_application <- as.factor(q1$fertilizer_application)
str(q1)
```


```{r}
plot(q1$plot)
```

The plot 160 should be labelled as 60, and the plot 60 should be labelled as 160.   

```{r, warning = F}
desplot::desplot(
  q1,
  rep ~ col * row,
  cex = 1,
  text = cultivar,
  out1 = cultivar
)
```

The row 5 with columns 21 to 25 should be labelled as rep 3 rather than rep 5.     
The row 8 with columns 1 to 5 should be labelled as rep 3 rather than rep 3, otherwise cultivar 20 would be present twice within rep 4.    
The row 10 with columns 20 to 24 should be labelled as rep 5 rather than rep 2.   
Fixing this, now each cultivar appears only once in each rep.   

__2.	Design an experiment to answer the following questions (Attach the CSV file__
__and include below a figure with the layout of your experiment):__  

a) Does increasing the dose of nitrogen affect yield? Does it depend on the 
frequency of irrigation? or on the cultivar? In the allocated field, I have some areas
that are more fertile than others. I have resources for a total of 90 EU.  

b) Is there a variation in bacteria/fungus growth with different agar types 
(semi-solid, gel-like state)? I have 30 plates (experimental unit) available.

__3.The data set `mcconway.turnip` from the package `agridat` presents us with an RCBD__
__experiment of turnips with 16 treatments allocated at random to each of four blocks __
__The 16 treatments were combinations of two varieties, two planting dates, and four densities.__
a) Run anova as usual. Are the requirements met?
b) Run a mixed model considering block as a random term. 
c) If there is any issue with the data that result in not meeting the ANOVA 
assumptions, use a mixed model to solve it.

```{r, message=FALSE}
library(tidyverse)
q3 <- agridat::mcconway.turnip %>%
  as_tibble() %>%
  mutate(densf = density %>% as.factor)
```

